### 2025-08-19 cognitive-bias.qmd
20250819 - updates

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Thus, a "low base-rate" event is an event with a low probability of occurring (i.e., an unlikely event).
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
This is especially true if the unlikely event attracts attention, because of the [confirmatory bias](#sec-cognitiveBiasesConfirmation) of memory [@Kahneman2011].

@@

The [framing](#sec-cognitiveBiasesFraming) of an event can also alter its weighting.
For instance, concrete representations tend to be weighted more heavily than abstract representations [@Kahneman2011].
Abstract representations are framed in terms of "how likely" the event will occur in in terms of chance, risk, probability, or likelihood.
Concrete representations are framed in terms of  "how many" people will be affected by the event.
An example abstract representation is that there is a 0.001% chance of dying from this particular disease.
When framed in this way, the risk appears small.
Now, consider the same information framed in terms of a concrete representation: 1 in 100,000 people will die from this disease.
Due to a phenomenon called *denominator neglect*, low base-rate events are more heavily weighted when they are framed using a concrete representation than an abstract representation [@Kahneman2011].

>>>>

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Thus, a "low base-rate" event is an event with a low probability of occurring (i.e., an unlikely event).
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
This is especially true if the unlikely event attracts attention, because of the [confirmatory bias](#sec-cognitiveBiasesConfirmation) of memory—that is, when thinking about the event, you try to make it true in your mind [@Kahneman2011].

@@

The [framing](#sec-cognitiveBiasesFraming) of an event can also alter its [availability](#sec-heuristicsAvailability) in mind, judgments of probability, and weighting in decisions [@Kahneman2011].
A more vivid description of an outcome can lead people to be more influenced by its ease of recall (i.e., [availability](#sec-heuristicsAvailability)) than its objective probability [@Kahneman2011].
In addition, concrete representations tend to be weighted more heavily than abstract representations [@Kahneman2011].
Abstract representations are framed in terms of "how likely" the event will occur in in terms of chance, risk, probability, or likelihood.
Concrete representations are framed in terms of  "how many" people will be affected by the event.
An example abstract representation is that there is a 0.001% chance of dying from this particular disease.
When framed in this way, the risk appears small.
Now, consider the same information framed in terms of a concrete representation: 1 in 100,000 people will die from this disease.
Due to a phenomenon called *denominator neglect*, low base-rate events are more heavily weighted when they are framed using a concrete representation than an abstract representation [@Kahneman2011].
